---
title: "Homework 2"
author: "Ekaterina Hofrenning"
date: '2023-09-27'
output: github_document
---

```{r}
library(tidyverse)
library(lubridate)
library(readxl)
```

# Problem 1

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols <- 
  read.csv("./data/fivethirtyeight_datasets/pols-month.csv") %>%  #read pols-month data
  separate(mon, sep = "-", into = c("year", "month", "day")) %>% # y/m/d variables
  mutate_at(c("year", "month", "day"), as.numeric) %>% # convert to numeric
  mutate(month = month.abb[month], # jan-dec
         president = case_when(prez_gop == 1 ~ "gop", prez_dem == 1 ~ "dem")) %>% # combine into president
  select(-c(prez_dem, prez_gop, day))
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp <- 
  read.csv("./data/fivethirtyeight_datasets/snp.csv") %>%
  separate(date, sep = "/", into = c("month", "day", "year")) %>%  # m/d/y variables
  mutate_at(c("year", "month", "day"), as.numeric) %>% # convert to numeric
  mutate(month = month.abb[month]) %>% # jan-dec
  select(-day) %>% # remove day
  relocate(year, month)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
unemployment <- 
  read.csv("./data/fivethirtyeight_datasets/unemployment.csv") %>%
  pivot_longer(names_to = "month", cols = Jan:Dec) %>% # switch to long format
  rename(year = Year) # match format
```

Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
merge1 <- left_join(pols, snp, by = c("year", "month"))
merge2 <- left_join(merge1, unemployment, by = c("year", "month"))
```


Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

```{r}
dim(merge2)
names(merge2)
min(merge2$year)
max(merge2$year)
```

The resulting dataset contains 11 columns and 822 observations, containing data from 1947 to 2015. Some key varibles include: part of current president, number of GOP senators, number of Democratic senators, among others.

# Problem 2

Read and clean the Mr. Trash Wheel sheet:

- specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
- use reasonable variable names
- omit rows that do not include dumpster-specific data

The data include a column for the (approximate) number of homes powered. This calculation is described in the Homes powered note, but not applied to every row in the dataset. Update the data to include a new homes_powered variable based on this calculation.

```{r}
mrtrash <- read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = 1) %>%
  filter(!is.na(Dumpster)) %>%
  mutate(Date = as.Date(Date))
```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?






